---
title: "mercat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mercat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mercat, quietly = TRUE)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
```


## Overview

MERCAT is the Molecular Epidemiology Researcher's Collection of Analytical Tools.
These tools are designed to analyse and visualise molecular data 
  from bacterial isolates.
For example it can be used to investigate spoligotypes and multi-locus VNTR data 
  from _Mycobacterium tuberculosis_. It can also be used to visualise multiple drug resistance data.

MERCAT expands and improves on an earlier online utility called spolTools, 
which was limited to spoligotypes from _M. tuberculosis_.
Improvements include:

  * visualisation of a range of genetic markers (eg. spoligotypes, VNTRs) from any bacteria
  * incorporation of antibiotic resistance data
  * a minimum spanning tree depicting relationships among isolates
  * a variety of frequency plots.

Please note that the package name for MERCAT is lowercase `mercat` but we refer to the software as MERCAT.

## License

MERCAT is licensed under the GNU LGPL-3.0 License. This license is included with
the package and you can view it with the following command:

```
file.show(system.file("extdata/lgpl-3.0.txt", package = "mercat"))
```

## Getting started with MERCAT

To list all functions available in MERCAT use 

```
help(package = "mercat")
```

Throughout this vignette we will use a preloaded data file containing genotype data sampled 
from tuberculosis found in 100 patients in Bolivia, 
[Genotypes of Mycobacterium tuberculosis in patients at risk of drug resistance in Bolivia](http://www.ncbi.nlm.nih.gov/pubmed/23603419) published in 2013.
The name of the example R data object is `monteserin`:

```
help(monteserin)
```

The class of this object is `merdata`:

```{r}
class(monteserin)
```

And you can use the S3 methods `print` and `summary` on a `merdata` object:

```
print(monteserin)
summary(monteserin)
```

## MERCAT interactive

The MERCAT R package contains a web interface written in Shiny and JavaScript.
This interface showcases several of the MERCAT plots, allows you to upload your
own data and has interactive D3 based network plots for a spoligoforest and minimium
spanning tree (MST). 

To try out MERCAT interactive type:

```
m_interactive()
```

## File import and export and merdata creation

MERCAT supports file read and export with the Rich Spoligotype Format (RSF) and 
the Simple Molecular Epidemological Format (SMEF). 

(for RSF c.f. spolTools: online utilities for analyzing spoligotypes of the 
_Mycobacterium tuberculosis_ complex. _Bioinformatics_.)

Additionally you can create a `merdata` object inside of R.

### Saving and loading SMEF and RSF files

You can load from `SMEF` and `RSF` files and create `merdata` objects by using the
`m_read` function:

```
m_read("Monteserin.smef")
```


You can save a `merdata` object to a file by using the `m_export` function. 

```
m_export(monteserin, "monteserin.smef")
```

We include a number of example SMEF and RSF files with the package. You can list them with:
```{r}
list.files(system.file("extdata/datasets", package = "mercat"))
```

And you can view a file (e.g. `Douglas.rsf`) with
```
file.show(system.file("extdata/datasets/Douglas.rsf", package = "mercat"))
```

### Creating a merdata object in R from two data frames

You can use the function `m_create` to make a `merdata` object from two data frames. 
One data frame contains `details` about the the `merdata` while the other data frame
contains the marker `data`:

```{r}
m_create(
  details = data.frame(Format = "SMEF 1.0", Author = "John Smith", pubyear = 1900), 
  data = data.frame(
  ID = c("1", "2", "3", "4", "5", "6", "7", "8"),
  COUNT = c("5", "2", "3", "7", "13", "2", "9", "4"),
  MLVA = c("25354323", "25753323", "25353322", "25353443",
           "25553323", "25353923", "25353323", "28353323"),
  NOTE = c("1", "1", "6", "4", "5", "6", "7", "7")
))
```

## Aggregating merdata and summary statistics

One of the powerful features of MERCAT is the ability to quickly aggregate
molecular data from different sources. As an example we can print the `summary` 
for the entire `monteserin` merdata object:

```{r}
summary(monteserin)$FULL
```

We know these data contain both `SPOL` and `MVLA15` types. We would like to
aggregate so only `SPOL` data is considered:

```{r}
summary(m_aggregate(monteserin, "SPOL"))$FULL
```

The `summary` function also has marker specific summary statistic output:

```{r}
summary(monteserin)$SPOL$spacer_freq
```

The entire `summary` output is quite lengthy but contains useful information. The
general way to access this information is `summary(<file>)$<MARKER>$<statistic>`.

## Plotting merdata

MERCAT includes a number of plotting functions that all use `ggplot2`. Each of
the functions will return a `ggplot2` object that can be manipulated further 
using standard `ggplot2` commands.

### Resistance frequencies

The `m_res_freq` function returns a bar graph showing the frequency of antibiotic
drug resistance in the `merdata` object:

```{r, fig.width = 7, fig.height = 4}
m_res_freq(monteserin, orderbynames = FALSE, reverseorder = TRUE)
```

### Spoligotype frequencies

You can easily make a nice spoligotype frequency plot with `m_spol_freq`:

```{r, fig.width = 7, fig.height = 8}
m_spol_freq(monteserin, label_size = 8)
```

### MLVA frequencies

All of the plotting functions allow graphical customisation. Here we hide
the values on a MLVA frequency plot and add a legend. We also change the 
colour of the tiles:

```{r, fig.width = 7, fig.height = 4.5}
m_mlva_freq(monteserin, col_high = "darkgreen", values = FALSE, legend = TRUE)
```

### Antibiotic Drug Resistance Correlation

You can plot the association between antibiotic drugs with respect to frequency
with the function `m_res_corr_mat`. This borrows from the population genetics
quantification of linkage disequilibrium. For further details see:

Slatkin (2008). Linkage disequilibrium-understanding the evolutionary past and mapping the medical future. _Nature Reviews Genetics_, 9(6):477-485, 2008.

```{r, fig.width = 5, fig.height = 4}
m_res_corr_mat(monteserin, text_size = 5)
m_antibiotic_full_names(c("STM", "INH", "RMP", "EMB", "PZA"))
```

## Visualising genotype and resistance data

MERCAT includes two network plots for visualising genotype and resistance data: 
`m_spolforest` and `m_mst`. Both of the plots use an underlying plotting function
called `m_plot_graph`. Many graphical options can be configured for these plots
and looking at the help file for `m_plot_graph` will give you an idea of how to
tailor the plot to suit your needs.

There are two functions that accompany the plots which provide detailed information
about the nodes shown in the plots. These functions are: `m_spolforest_nodes` and
`m_mst_nodes`. The underlying function for these is called `m_print_nodes`. 

### Spoligoforests

You can produce a spoligoforest combined with drug resistance genotypes:

```{r, fig.width = 7, fig.height = 3} 
m_spolforest(monteserin, node_label_size = 3, tree_seed = 4)
```

To get a data frame containing the corresponding node data use `m_spolforest_nodes` with the same
tree_seed value:

```
m_spolforest_nodes(monteserin, tree_seed = 4)
```

If you would prefer a spoligoforest without drug resistance genotypes you can easily
aggregate the `merdata` prior to plotting:

```{r, fig.width = 7, fig.height = 3} 
m_spolforest(monteserin %>% m_aggregate("SPOL"), node_label_size = 3, tree_seed = 4)
```

### Minimum Spanning Trees

Similiar to the spoligoforest you can plot a minimum spanning tree:

```{r, fig.width = 7, fig.height = 7}
m_mst(monteserin, tree_seed = 6, plot_scale_x = 0.6, plot_scale_y = 0.6)
```

To get the data frame with corresponding node data use `m_mst_nodes` with the same
tree_seed value:

```
m_mst_nodes(monteserin, tree_seed = 6)
```

## Credits
The MERCAT team is RZ Aandahl, N Vaudagnotto, S Bhatia, AG Street, AR Francis, and MM Tanaka. 

MERCAT interactive was developed by AG Street and RZ Aandahl.

spolTools was designed by C Tang, JF Reyes, F Luciani, AR Francis, and MM Tanaka.

## References
Aandahl RZ, Bhatia S, Vaudagnotto N, Street AG, Francis AR, Tanaka MM. MERCAT: Visualising molecular epidemiology data combining genetic markers and drug resistance profiles. In preparation.

Reyes JF, Francis AR, Tanaka MM. Models of deletion for visualizing bacterial variation: an application to tuberculosis spoligotypes. _BMC Bioinformatics_. 2008 Nov 27;9:496. doi: 10.1186/1471-2105-9-496.

Tang C, Reyes JF, Luciani F, Francis AR, and Tanaka MM. spolTools: online utilities for analyzing spoligotypes of the _Mycobacterium tuberculosis_ complex. _Bioinformatics_. 2008 Oct 15;24(20):2414-5. Epub 2008 Aug 18.